Do LSTMs Understand the Licensing of Negative Polarity Items?

### 2021-12-24 paper/paper.tex

There is, though, little research investigating the feature of licensing of negative polarity items in language. Negative polarity items are a class of words that are only permitted to appear in certain "negative" contexts. For example, in English, negative polarity items include words like \textit{ever}, \textit{any}, and \textit{even}.
\begin{exe}
\ex \begin{xlist}
\ex \label{ex:example1} The boy did not ever give up.
\ex[*]{\label{ex:example2} The boy did ever give up.}
\ex[*]{\label{ex:example3} The boy who did not win did ever give up.}
\end{xlist}
\end{exe}
Considering Examples \ref{ex:example1} and \ref{ex:example2}, we note that the word \textit{ever} is only permitted to occur in a sentence when it is licensed by a word such as \textit{nobody} which provides a ``negative'' context. For this basic example, the relationhip between licensor and NPI would not be a particularly challenging task for a language model to learn; if it doesn't see any negative words in environment, then it won't predict an NPI.

@@

If a language model after being trained on a generic linguistic data set is able to predict the distribution of negative polarity items accurately, this would be evidence that the model is able to learn this feature of syntax, and in particular, that it represents hierarchical structure with enough fidelity to capture the c-command relation.

@@

For our investigation, we trained a LSTM model on the WikiText-2 dataset \cite{merity2016pointer}, which consists of over 2 million tokens from 600 Wikipedia articles, with a vocabulary size of 33,279. Our LSTM has an embedding layer, assigning 200-dimensional embeddings to each word in the vocabulary, and two recurrent layers, each with 650 hidden units. During training, we use dropout regularization as described in Srivastava et. al. \shortcite{srivastava2014dropout}, with a dropout rate set at 0.2. Optimization was performed using gradient descent, with a base learning rate of 0.2, which decreased exponentially whenever performance on the development set regressed.

The model can be found in the project's GitHub repository (\url{https://github.com/tsandstr/ling380-final-project}).

@@

A high surprisal metric means that, in the given context, a word was predicted with \textit{low} probability, and was therefore unexpected by the model. Basically, given a hidden state of the model at a current time step, the surprisal metric is a way to measure the whether the next word makes \textit{sense}. A direct correlation has been shown between language model surprisal metrics and  human sentence processing difficulty \cite{hale2001probabilistic,levy2008expectation,smith2013effect}.

\subsection{Experimental Design}

@@

Negative polarity items have a variety of linguistic properties which we would like to probe. On the most basic level, they are only allowed to occur in negative contexts. More specifically, in order for an NPI to be licit, it must be c-commanded by its licensor. Another property is that NPIs are immune to intervening material separating the NPI from its licensor. And finally, one negative context can license multiple negative polarity items. In this section we demonstrate that LSTMs have learned these four properties negative polarity items.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{surprisal-c-command-box-plot}
    \caption{Plots of surprisal scores in the presence of a licensor and without a licensor. As expected, surprisal tends to be higher in the environment with no licensor. Interstingly, \textit{median} instantaneous surprisal actually decreases when we include a licensor. However, this does not take into account the fact that our dataset is paired; this is only captured when we apply a $t$-test to the licensing interaction.}
    \label{fig:basic-detection-box-plot}
\end{figure}

@@

@@

\subsection{Robustness to Intervening Material}

Because the relation between licensor and NPI is based on the c-command relation rather than on linear word order, it is not sensitive to the amount of intervening material which occurs between the licensor and the NPI. In Example \ref{ex:intervening}, the NPI \textit{anymore} is licenced by the word \textit{not} in each sentence. The intervening prepositional phrases and relative clauses have no impact on the grammaticality of the sentences.
\begin{exe}
\ex \label{ex:intervening}
\begin{xlist}
\ex We do not go to the supermarket anymore.
\ex We do not go to the supermarket that is owned by Mark anymore.
\ex We do not go down the block to the supermarket that is owned by Mark anymore. 
\end{xlist}
\end{exe}
We showed in Section \ref{sec:basic} that LSTMs exhibit significantly higher surprisal when encountering an unlicensed NPI than a licensed NPI; however, the analysis in Section \ref{sec:c-command} suggests that the model is using a heuristic based on linear word order rather than taking into account the hierarchical structure of the sentence. Given this, we might predict that the amount of intervening material could have an impact on surprisal.

To test the robustness of the model's predictions to intervening material, we first designed 21 test items containing NPIs. In constructing these items, we made sure to vary the amount of intervening material between the licensor and the NPI. Then, for each test item, we constructed two variations: one with a licensor and one without.

We also did a slightly different experiment aiming to control for variation in the semantic content of the sentences. Here, we designed 10 sentences base sentences with NPI items. Each of the base sentences comes in three varieties: minimal intervening material (enough to make the sentence grammatical), 3-5 words of additional intervening material, and 6-10 additional words between the negative licensor and the polarity item. In all cases the intervening words modify the object of the base sentence. An example of the results for this section are visualized in Figure 2.

\begin{figure*}
    \centering
    \subfloat[Base Sentence]{{\includegraphics[width=.32\textwidth]{short} }}
    \subfloat[5 Additional Words]{{\includegraphics[width=.32\textwidth]{medium} }}
    \subfloat[8 Additional Words]{{\includegraphics[width=.32\textwidth]{long} }}
    \caption{Plots of surprisal scores for words in the modified sentence. The surprisal score of the negative polarity item \textit{anymore} remains relatively unchanged by the amount of intervening material.}
\end{figure*}

We once again found that there was a weak positive correlation between amount of intervening material and the licensing interaction. However as seen in the plots, the scores are relatively similar. This test supports the notation that the model's ability to predict the the distribution of NPIs is not affected by the quantity of intervening material.

\subsection{Ability to License Multiple Negative Polarity Items}
In most cases, a negative context can license multiple negative polarity items.
\begin{exe}
\ex
\begin{xlist}
\ex He didn't want to go with anybody anywhere
\ex [*]{He did want to go with anybody anywhere} 
\end{xlist}
\end{exe}

\section{Results}

>>>>

There is, though, little research investigating the feature of licensing of negative polarity items in language. Negative polarity items are a class of words that are only permitted to appear in certain ``negative'' contexts. For example, in English, negative polarity items include words like \textit{ever}, \textit{any}, and \textit{even}.
\begin{exe}
\ex \begin{xlist}
\ex \label{ex:example1} The boy did not ever give up.
\ex[*]{\label{ex:example2} The boy did ever give up.}
\ex[*]{\label{ex:example3} The boy who did not win did ever give up.}
\end{xlist}
\end{exe}
Considering Examples \ref{ex:example1} and \ref{ex:example2}, we note that the word \textit{ever} is only permitted to occur in a sentence when it is licensed by a word such as \textit{nobody} which provides a ``negative'' context. For this basic example, the relationhip between licensor and NPI would not be a particularly challenging task for a language model to learn; if it doesn't see any negative words in environment, then it won't predict an NPI.

@@

If a language model trained on a generic linguistic data set is able to predict the distribution of negative polarity items accurately, this would be evidence that the model is able to learn this feature of syntax, and in particular, that its internal representation captures hierarchical structure with enough fidelity to encapsulate the c-command relation.

@@

For our investigation, we trained a LSTM model on the WikiText-2 data set \cite{merity2016pointer}, which consists of over 2 million tokens from 600 Wikipedia articles, with a vocabulary size of 33,279. Our LSTM has an embedding layer, assigning 200-dimensional embeddings to each word in the vocabulary, and two recurrent layers, each with 650 hidden units. During training, we use dropout regularization as described in Srivastava et. al. \shortcite{srivastava2014dropout}, with a dropout rate set at 0.2. Optimization was performed using gradient descent, with a base learning rate of 0.2, which decreased exponentially whenever performance on the development set regressed.

The model can be found in the project's GitHub repository\footnote{\url{https://github.com/tsandstr/ling380-final-project}}.

@@

A high surprisal metric means that, in the given context, a word was predicted with \textit{low} probability, and was therefore unexpected by the model. Intuitively, given a hidden state of the model at a current time step, the surprisal metric is a way to measure the whether the next word makes \textit{sense}. A direct correlation has been shown between language model surprisal metrics and  human sentence processing difficulty \cite{hale2001probabilistic,levy2008expectation,smith2013effect}.

@@

Negative polarity items have a variety of linguistic properties which we would like to probe. On the most basic level, they are only allowed to occur in negative contexts. More specifically, in order for an NPI to be licit, it must be c-commanded by its licensor. Another property is that NPIs are immune to intervening material separating the NPI from its licensor. In this section we  investigate whether LSTMs understand these three properties.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.5\textwidth]{surprisal-c-command-box-plot}
    \caption{Plots of surprisal scores in the presence of a licensor and without a licensor. As expected, surprisal tends to be higher in the environment with no licensor. Interstingly, \textit{median} instantaneous surprisal actually decreases when we include a licensor. However, this does not take into account the fact that our data set is paired; this is only captured when we apply a $t$-test to the licensing interaction.}
    \label{fig:basic-detection-box-plot}
\end{figure}

@@

Comparing the two statistically significant effects, we used partial-$\eta^2$ to measure the size of these effects (the (partial)-$\eta^2$ statistic for ANOVA is analogous to the $r^2$ statistic for linear regression, in that it described the proportion of variance in the dependent variable explained by each independent variable). We found that the effect size for the main licensor is $\eta_p^2 = 0.36$ and the effect size for the embedded licensor is $\eta_p^2 = 0.37$. Thus, both the main licensor and the embedded licensor have similar sized effets on instantaneous surprisal.

The fact that the interaction term is not significant means that the effect of the embedded licensor does not depend on the presence or lack of a main licensor. In other words, the effects of the licensors combine linearly.

@@

\subsection{Robustness to Intervening Material}
\label{sec:intervening}
Because the relation between licensor and NPI is based on the c-command relation rather than on linear word order, it is not sensitive to the amount of intervening material which occurs between the licensor and the NPI. In Example \ref{ex:intervening}, the NPI \textit{anymore} is licenced by the word \textit{not} in each sentence. The intervening prepositional phrases and relative clauses have no impact on the grammaticality of the sentences.
\begin{exe}
\ex \label{ex:intervening}
\begin{xlist}
\ex We do not go to the supermarket anymore.
\ex We do not go to the supermarket that is owned by Mark anymore.
\ex We do not go down the block to the supermarket that is owned by Mark anymore. 
\end{xlist}
\end{exe}
We showed in Section \ref{sec:basic} that LSTMs exhibit significantly higher surprisal when encountering an unlicensed NPI than a licensed NPI; however, the analysis in Section \ref{sec:c-command} suggests that the model is using a heuristic based on linear word order rather than taking into account the hierarchical structure of the sentence. Given this, we might predict that the amount of intervening material could have an impact on surprisal.

To test the robustness of the model's predictions to intervening material, we first designed 21 test items containing NPIs. In constructing these items, we made sure to vary the amount of intervening material between the licensor and the NPI. Then, for each test item, we constructed two variations: one with a licensor and one without. For each item, we computed the licensing interaction, and then checked for a correlation between the amount of intervening material and the licensing interaction. Our results are visualized in Figure \ref{fig:intervening1}. We found no statistically signficant correlation between intervening length and either instantaneous licensing interaction ($p = 0.305$) or cumulative licensing interaction ($p = 0.485$). This suggests that the model's ability to predict NPI distribution is not affected by the mount of intervening material between the NPI and its licensor.

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{intervening_pnt_scatter}
        \caption{Instantaneous licensing interaction}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{0.45\textwidth}
        \centering
        \includegraphics[width=\textwidth]{intervening_cum_scatter}
        \caption{Cumulative licensing interaction}
    \end{subfigure}
    \caption{Plots of licensing interaction of sentences given the amount of words in between the negative licensor and the NPI}
    \label{fig:intervening1}
\end{figure}
\begin{figure*}
    \centering
    \subfloat[Base Sentence]{{\includegraphics[width=.32\textwidth]{short} }}
    \subfloat[5 Additional Words]{{\includegraphics[width=.32\textwidth]{medium} }}
    \subfloat[8 Additional Words]{{\includegraphics[width=.32\textwidth]{long} }}
    \caption{Plots of surprisal scores for words in the modified sentence. The surprisal score of the negative polarity item \textit{anymore} remains relatively unchanged by the amount of intervening material.}
    \label{fig:intervening2}
\end{figure*}
We also did a slightly different experiment aiming to control for variation in the semantic content of the sentences. Here, we designed 10 sentences base sentences with NPI items. Each of the base sentences comes in three varieties: minimal intervening material (enough to make the sentence grammatical), 3-5 words of additional intervening material, and 6-10 additional words between the negative licensor and the polarity item. In all cases the intervening words modify the object of the base sentence. An example of the results for this section are visualized in Figure \ref{fig:intervening2}.

We once again found that there was a weak positive correlation between amount of intervening material and the licensing interaction. However, as seen in the plots, the scores are relatively similar. This test supports the notation that the model's ability to predict the the distribution of NPIs is not affected by the quantity of intervening material.

% \subsection{Ability to License Multiple Negative Polarity Items}
% In most cases, a negative context can license multiple negative polarity items.
% \begin{exe}
% \ex
% \begin{xlist}
% \ex He didn't want to go with anybody anywhere
% \ex [*]{He did want to go with anybody anywhere} 
% \end{xlist}
% \end{exe} 
\section{Positive Polarity Items}
We briefly turn our attention to positive polarity items (PPIs) which are words that can only occur in ``positive'' contexts. English words such as \textit{some, rather,} and \textit{already} are common examples of PPIs. There is significantly less literature discussing PPIs than NPIs, and often the licensing of a positive context is much less explicit than that of a negative context. We are interested in whether LSTMs will exhibit higher surprisal in unlicensed settings, and if so, does the model understand PPIs to the same extent as NPIs. Establishing a positive context in English is not as explicit as establishing a negative context. Sentences default to the positive setting, and thus there is not an equivalent to \textit{not} which licenses a negative context. We hypothesize that the lack of an obvious licensor might make it more difficult for the model to determine whether a PPI is permissible.

\subsection{Licensing of Positive Polarity Items}
Positive polarity items may only be licensed in positive contexts, as is shown for the PPI \textit{rather} in the following example.
\begin{exe}
\ex \label{ex:positive}
\begin{xlist}
\ex I would rather be there
\ex[*]{I wouldn't rather be there}

\end{xlist}
\end{exe}
Plotted in Figure \ref{fig:ppi-basic-licensing} are graphs surprisal of the sentences in \ref{ex:positive}. One can see that the positive polarity item has a slightly lower surprisal score in the grammatically correct context, but the trend is not as pronounced as what was seen with NPIs.

We conduct a nearly identical experiment as that in Section \ref{sec:basic} to determine the extent to which LSTMs learn about PPIs. We constructed 10 test items containing positive polarity items, and for each item, we constructed one with a positive context, and another with a negative context as exemplified in Example \ref{ex:positive}. If the model learned to predict PPIs only in positive contexts, then we would expect the licensing interaction of these sentences to be positive.

\begin{figure}
    \centering
    \subfloat[Grammatical]{{\includegraphics[width=.42\linewidth]{img/ppigood.png} }}
    \subfloat[Ungrammatical]{{\includegraphics[width=.42\linewidth]{img/ppibad.png} }}
     \caption{The surprisal score for \textit{rather} is slightly lower in the positive context than in the incorrect negative context}
    \label{fig:ppi-basic-licensing}
\end{figure}

Measuring both instantaneous surprisal and cumulative surprisal after a PPI, we found in both cases that the mean licensing interaction is slightly greater than 0 (which implies that surprisal is higher when we find a PPI in a negative context), but the difference is not large enough to be statistically significant (instantaneous, $t(9) = 1.43$, $p = 0.093$; cumulative, $t(9) = 1.37$, $p = 0.102$). Thus, we fail to reject the null hypothesis that the licensing interaction is zero.

This small experiment suggests that it is more difficult for an LSTM to determine the validity of a PPI than an NPI. This result is not particularly surprising because, unlike NPIs, PPIs lack explicit licensors. Thus, in order to correctly predict the distribution of PPIs, the model would need to represent more complex aspects of language.

\section{Results and Discussion}

In Section \ref{sec:basic}, we showed that the presence of a negative licensor in a sentence caused a significant decrease in the model's surprisal upon encountering an NPI. This suggests that the model has learned that NPIs only occur when a negative licensor is present. However, in Section \ref{sec:c-command}, we saw that the model does not distinguish between licensors which c-command the NPI and those that do not. This suggests that, rather than representing the hierarchical structure of the sentence, the model approximates the NPI/licensor relation using a representation based on linear word order and precedence. Finally, in Section \ref{sec:intervening}, we saw that the model's ability to predict the distribution of NPIs was independent of the amount of intervening material.

Despite the fact that the experiments in Section \ref{sec:c-command} suggested that the model does not represent the c-command relation, the results do not differ too dramatically from what is observed in psycholinguistic studies of NPI licensing. As we would expect, listeners with no time constraints tend to judge sentences like those in Examples \ref{ex:c-command-np} and \ref{ex:c-command-pp} as being ungrammatical. However, when fast time constraints are placed on the judgement task, some speakers mistakenly accept sentences like Example \ref{ex:c-command-np} (in which there exists a negative licensor, but it is contained in an embedded clause) as grammatical \cite{xiang2009illusory,xiang2013dependency,parker2016negative}. This suggests that, under time pressure, the human brain might process the NPI licensing relation using a heuristic similar to the one that our model has learned, which takes into account only precedence and not c-command.

These results point towards a number of questions which could be addressed in future research. First of all, we might like to know how well these results generalize to other models. Our homegrown model was trained on a relatively small data set using limited resources; perhaps a larger LSTM like that presented in Jozefowicz et. al. \shortcite{jozefowicz2016exploring} would have a better chance of correctly modeling the c-command relationship. We could also apply a similar analysis to other types of network architectures, such as transformers. Large transformers like the GPT-2 model presented by Radford et. al. \shortcite{radford2019language} and the BERT model introduced in Devlin et. al. \shortcite{devlin2018bert} have both performed very well on a very general set of language modeling tasks. Repeating a similar analysis on GPT-2 or Bert could provide interesting insights into the extent to which transformer models represent hierarchical structure.

Of course, there is also further research that could be done without moving to a new model or architecture. One possibility would be to try to identify a specific unit or collection of units in the cell state of the LSTM that is responsible for tracking negative and positive states. This sort of analysis has been achieved before; in Lakretz et. al. \shortcite{lakretz2019emergence}, the authors identified discrete units responsible for tracking information about number agreement. They achieved this by using a technique known as \textit{ablation}. First, they constructed a task which measured the models ability to track number agreement. Then, one by one, they zeroed out (ablated) individual units in the LSTMs cell state. By selecting units whose ablation was associated with decreased success on the number agreement task, they were able to identify specific units responsible for tracking number agreement.

If a similar ablative analysis could identify specific units responsible for tracking negative polarity licensing, we could make predictions about how this unit's activation would change over the course of a sentence. In Section \ref{sec:c-command}, we observed that both embedded licensors and main licensors have a significant effect on surprisal, and that these effects combine linearly. This suggests that a hypothetical ``NPI licensing unit'' in the LSTM's hidden state should be expected to increase in activation each time the model encounters a negative word, regardless of the current level of activation. If we could observe this pattern in a specific unit of the LSTM's cell state, this would confirm that the unit identified by the ablative analysis was in fact responsible for tracking NPI licensing.
